Harnessing generative AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology

### 2024-05-22 docs/_tex/manuscript.tex
Add fig-congenital-branches

>>>>

\subsubsection{Congenital onset by HPO
branch}\label{congenital-onset-by-hpo-branch}

\phantomsection\label{cell-fig-congenital-branches}
\begin{figure}[H]

\includegraphics{manuscript_files/figure-pdf/fig-congenital-branches-1.pdf}

\caption{\label{fig-congenital-branches}Distribution of congenital onset
across HPO branches. The y-axis shows the proportion of phenotypes that
are always/often/rarely/never congenital. The x-axis shows the HPO
branch, orderered from highest to lowest proportion of always congenital
phenotypes.}

Next, we assessed the distribution of congenital onset across HPO
branches (Fig.~\ref{fig-congenital-branches}). We found that the
Abnormality of prenatal development or birth branch contained the
greatest proportion of phenotypes that were always congenital (70.15\%),
followed by Abnormality of the musculoskeletal system (45.34\%) and
Growth abnormality (37.62\%). This is concordant with the expectation
that these phenotypes should largely be congenital. The HPO branches
with the least commonly congenital phenotypes were Constitutional
symptom (0\%), Abnormality of the thoracic cavity (0\%), and Phenotypic
abnormality (0\%).
\href{https://hpo.jax.org/app/browse/term/HP:0025142}{`Constitutional
symptom'} is a fairly broad term defined as \emph{`A symptom or
manifestation indicating a systemic or general effect of a disease and
that may affect the general well-being or status of an individual.'}
Examples include `Fatigue' `Exercise intolerance', `Hot flashes' and
`Sneeze'.

### 2024-05-22 docs/manuscript.qmd
Add fig-congenital-branches

>>>>

### Congenital onset by HPO branch

```{r gpt_annot_plot_branches}
gpt_annot_plot_branches_out <- HPOExplorer::gpt_annot_plot_branches(
  gpt_annot = gpt_annot, 
  hpo = hpo, 
  metric = "congenital_onset", 
  fill_lab = "Congenital\nonset", 
  show_plot = FALSE)
top_congenital <- unique(gpt_annot_plot_branches_out$data[!is.na(congenital_onset),c("ancestor_name","proportion_always")])
```

```{r fig-congenital-branches}
#| label: fig-congenital-branches
#| fig-cap: Distribution of congenital onset across HPO branches. The y-axis shows the proportion of phenotypes that are always/often/rarely/never congenital. The x-axis shows the HPO branch, orderered from highest to lowest proportion of always congenital phenotypes.
#| fig-height: 6
#| fig-width: 6

gpt_annot_plot_branches_out$plot
```

Next, we assessed the distribution of congenital onset across HPO branches ([Fig. @fig-congenital-branches]). We found that the `r top_congenital$ancestor_name[1]` branch contained the greatest proportion of phenotypes that were always congenital (`r round(top_congenital$proportion_always[1]*100,2)`%), followed by `r top_congenital$ancestor_name[2]` (`r round(top_congenital$proportion_always[2]*100,2)`%) and `r top_congenital$ancestor_name[3]` (`r round(top_congenital$proportion_always[3]*100,2)`%). This is concordant with the expectation that these phenotypes should largely be congenital. 
The HPO branches with the least commonly congenital phenotypes were `r rev(top_congenital$ancestor_name)[1]` (`r round(rev(top_congenital$proportion_always)[1]*100,2)`%), `r rev(top_congenital$ancestor_name)[2]` (`r round(rev(top_congenital$proportion_always)[2]*100,2)`%), and `r rev(top_congenital$ancestor_name)[3]` (`r round(rev(top_congenital$proportion_always)[3]*100,2)`%). ['Constitutional symptom'](https://hpo.jax.org/app/browse/term/HP:0025142) is a fairly broad term defined as *'A symptom or manifestation indicating a systemic or general effect of a disease and that may affect the general well-being or status of an individual.'* Examples include 'Fatigue' 'Exercise intolerance', 'Hot flashes' and 'Sneeze'.
